---
title: GSA Stipend & Cost-of-Living Analysis
subtitle: Appendix B. Cost of Living Analysis
author: Alex Barth, Secretary of Data Management
date: "`r Sys.Date()`"
format:
  pdf
execute:
  echo: FALSE
  warning: FALSE
  message: FALSE
---

```{r}
rm(list = ls())
library(ggplot2)
library(ggpubr)
library(ggusc)
library(dplyr)
library(HDInterval)


col_data <- read.table('../data/01_cost-of-living.tsv', sep= '\t')

```


# Cost of Living

### Distribution of Common Costs

To understand the cost of basic necessities of graduate students, the GSA cost of living survey asked respondents to indicate their basic expenditures in pre-assigned ranges. To quantify responses, values were recorded as the midpoint of the assigned range. Basic necessities include: Rent, Utilities, Food, Transportation, Health out-of-pocket, household goods. In total, 775 respondents were used in this analysis.

The average total cost of basic necessities was \$21,514.84. The high-density 50% interval was \$9,600 and \$34,800. It should be noted that the cost of basic necessities is below what is considered a living wage by the MIT living wage calculator.

```{r}

cat_to_numeric <- function(x) {
  
  if(any(x == '')){
    x[which(x=='')] <- '$0-$0'
  }
  
  x <-  strsplit(x, '-') |> 
    sapply(`[[`,1)
  
  #format as numeric
  x_num <- gsub("[^[:digit:]]", "", x) |> 
    as.numeric()
  
  #take those that are less than
  x_num[grep('Less',x)] <- 0
  
  
  x_num[is.na(x_num)] <- 0
  
  lower_bounds <- x_num |> unique() |> sort()
  
  mid_dist <- diff(lower_bounds)/2
  
  mid_values <- lower_bounds[1:length(lower_bounds)-1] + mid_dist
  mid_values <- c(mid_values, max(mid_values)+min(mid_dist*2))
  
  # assign mid values to previous full values
  
  x_out <- sapply(x_num,
                  function(x) mid_values[which(lower_bounds == x)])
  
}

col_data$rent_mo_num <- cat_to_numeric(col_data$rent_mo)
col_data$util_mo_num <- cat_to_numeric(col_data$util_mo)
col_data$food_mo_num <- cat_to_numeric(col_data$food_mo)
col_data$transp_mo_num <- cat_to_numeric(col_data$transp_mo)
col_data$healthCare_mo_num <- cat_to_numeric(col_data$healthCare_mo)
col_data$household_items_mo_num <- cat_to_numeric(col_data$household_items_mo)

# calculate total annual cost of living
col_data$housing_total_cost <- (col_data$rent_mo_num + col_data$util_mo_num) * 12
col_data$total_col <- apply(col_data[,grep('_num', names(col_data))],1,sum) * 12

```


## What is the average monthly housing cost?

The Average students spends \$`r round(mean(col_data$housing_total_cost/12),2)`/month on housing costs. However, several students report spending much more. `r round(1-ecdf(col_data$housing/12)(1200),4)*100`% of students reported paying over $1200.00/month.

```{r, fig.cap='Distribution of housing expenditures by graduate students.'}
library(geomtextpath)


# subtitle <- paste0('The avearge student spends $', round(mean(col_data$housing_total_cost/12),2),
#                    ' a month on housing expenses. However, several students pay much more. ',
#                    round(1-ecdf(col_data$housing_total_cost/12)(1200),4)*100, 
#                    '% of students reported paying over $1200/mo')


ggplot(col_data)+
  geom_density(aes(x = housing_total_cost / 12),
               fill = usc_cols(1)) +
  labs(x = 'Monthly Housing Expenses', y= '') +
  scale_y_continuous(expand = c(0,0),
                     labels = function(x) paste0(100*x, '%'))+
  theme_pubr()


# print(subtitle)
```


## What is the housing cost based on living situation?

Grad students live in several contexts and this can affect the total cost of housing.

```{r, fig.cap='Cost of housing based on living situation and the proportion of graduate students in each living situation.'}
library(treemapify)

col_data$living_situation[which(col_data$living_situation == '')] <- 'Other'
col_data$living_situation[grep('alone', col_data$living_situation)] <- 'Alone'
col_data$living_situation[grep('parents', col_data$living_situation)] <- 'Parents'
col_data$living_situation[grep('partner', col_data$living_situation)] <- 'Partner'
col_data$living_situation[grep('roommates', col_data$living_situation)] <- 'Roommates'

living_sit_sum <- col_data |> 
  group_by(living_situation) |> 
  summarize(mean_house = mean(housing_total_cost/12),
            sd_house = sd(housing_total_cost/12),
            count = n()) |> 
  mutate(freq = count/sum(count))

housing_cost_plot <- ggplot(living_sit_sum)+
  geom_bar(aes(x = living_situation, y = mean_house, fill = living_situation),
           stat = 'identity')+
  geom_errorbar(aes(x = living_situation, ymin = mean_house,
                    ymax = mean_house + sd_house, color = living_situation),
                width = 0.25) +
  labs(x = '', y = 'Monthly Housing Cost [$]')+
  scale_y_continuous(expand = c(0,0))+
  scale_fill_usc(palette = 'confident') +
  scale_color_usc(palette = 'confident')+
  guides(fill = 'none', color = 'none')+
  theme_pubr()

prop_plot <- ggplot(living_sit_sum,
                    aes(area = freq, fill = living_situation,
                        label = paste(living_situation, 
                                      paste0(round(freq,4)*100,'%'),
                                      sep = '\n')))+
  geom_treemap()+
  geom_treemap_text(color = 'white') +
  scale_fill_usc(palette = 'confident') +
  guides(fill = 'none')+
  theme_void()

ggarrange(housing_cost_plot, prop_plot, nrow = 2, heights = c(1,.5))
```

## What is the total cost of living distribution?



```{r, fig.cap='Distribution of total cost of basic necessities.'}

ggplot(col_data)+
  geom_density(aes(x = total_col),
               fill = usc_cols(1)) +
  labs(x = 'USD [$]', y= '',
       title = 'Annual Expenses') +
  scale_y_continuous(expand = c(0,0),
                     labels = function(x) paste0(100*x, '%'))+
  theme_pubr()


```


### How many people are reliant on external support?

Many students rely on external support from friends/family or will take out loans to help cover their expenses.
Overall, `r round(prop.table(table(col_data$take_loans))[['Yes']],4) * 100`% of respondents indicated that they took out loans. `r round(prop.table(table(col_data$recieve_fin_support)),4)[['Yes']] * 100`% of respondents stated they rely on friends or family for financial support. Those who receive financial support are less likely to take out aid.

```{r, fig.cap = 'Proportion of students reliant on external support and loans.'}
library(scales)

col_data <- col_data[col_data$recieve_fin_support != '',]
col_data <- col_data[col_data$take_loans != '',]

ggplot(col_data) +
  geom_bar(aes(x = recieve_fin_support, y = after_stat(count)/sum(after_stat(count)),
               fill = take_loans),
           position = 'stack')+
  labs(x = 'Dependent on family/friends financial support?',
       y = 'Percent', fill = 'Took out loans?')+
  scale_fill_usc(palette = 'confident')+
  scale_y_continuous(expand = c(0,0))+
  theme_pubr()+
  theme(legend.position = 'right')


```


For those who do not receive financial support from others, the average annual cost of living is `r round(mean(col_data$total_col[col_data$recieve_fin_support == 'No']),2)`. 

```{r}
ggplot(col_data)+
  geom_density(aes(x = total_col),
               fill = usc_cols(1)) +
  facet_wrap(~recieve_fin_support) +
  labs(x = 'USD [$]', y= '',
       subtitle = 'Separated by if students receive financial support',
       title = 'Annual Expenses') +
  scale_y_continuous(expand = c(0,0),
                     labels = function(x) paste0(100*x, '%'))+
  theme_pubr()
```
 

### How likely are people to take out loans based on stipend?

```{r}
stipends_clean <- read.table('../data/01_clean-stipends.tsv', sep = '\t')

col_stipends <- stipends_clean |> 
  select(RespondentId, stipend_amt, student_type, Program, program_type, College) |> 
  left_join(col_data, by = 'RespondentId')


col_stipends$take_loans <- col_stipends$take_loans |> 
  sapply(function(x) switch(x, 'Yes' = 1, 'No' = 0, NA)) |> 
  unlist()

loan_mod <- glm(take_loans ~ stipend_amt, family = 'binomial',
                data = col_stipends)

predicted_prob <- data.frame(
  stipend_amt = seq(0,35000,1000),
  probs = predict(loan_mod, 
                   newdata = data.frame(stipend_amt = seq(0,35000,1000)), 
                   type = 'response')
)

ggplot() +
  geom_point(aes(x = stipend_amt,
                 y = take_loans),
             data = col_stipends) +
  stat_smooth(data = predicted_prob,
             aes(x = stipend_amt,
                  y = probs),
             col = 'black', size = 1, method = 'gam')+
  labs(y = 'Probability of Taking Loan',
       x = 'Stipend [$]',
       title = 'How does stipend affect likelihood of taking out a loan?')+
  theme_pubr()

```


### How likely are people to take a loan based on degree type?

```{r}

```


Professional students are far more likely to take out loans and were excluded from this analysis. It was found that 43.1% of Master's students took loans while 21.1% of Doctoral students took loans.

```{r}
ggplot(col_stipends[!is.na(col_stipends$take_loans),]) +
  geom_bar(aes(x = student_type, y = after_stat(count)/sum(after_stat(count)),
               fill = ifelse(take_loans == 1, 'Yes','No')),
           position = 'stack')+
  labs(x = 'Student Type',
       y = 'Percent', fill = 'Took out loans?')+
  scale_fill_usc(palette = 'confident')+
  scale_y_continuous(expand = c(0,0))+
  theme_pubr()+
  theme(legend.position = 'right')



```

